Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-04-16 2015-eamt-melsa/2015-eamt-melsa.tex
Some stuff
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59737 72bbbca6-d526-0410-a7d9-f06f51895060
added the decipherment references, removed comments, and ironed out text a bit
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59858 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{abstract}
  This article presents a method of training maximum-entropy models to
  perform lexical selection in a rule-based machine translation
  system. The training method described is unsupervised; that is, it
  does not require any annotated corpus. The method uses
  source-language monolingual corpora, the machine translation (MT)
  system in which the models are integrated and a statistical
  target-language model. Using the MT system, the sentences in 
  the source-language corpus are
  translated in all possible ways according to the different
  translation equivalents in the bilingual dictionary of the system.
  These translations are then scored on the target-language model and
  the scores are normalised to provide fractional counts for training
  source-language maximum-entropy lexical-selection models. We show
  that these models can perform equally well, or better, than using
  the target-language model directly for lexical selection, at a
  substantially reduced computational cost.
\end{abstract}

\section{Introduction}
\todomlf{R1 has a general comment: ``watch out for long sentences''}
Corpus-based machine translation (MT) has been the primary research
direction in the field of MT in recent years. However, rule-based MT
(RBMT) systems are still being developed, and there are many
successful commercial and non-commercial systems. One reason for the
continued development of RBMT systems is that in order to be
successful, corpus-based MT requires parallel corpora in the order of
tens of millions of words. Although for some language pairs these
exist, they only exist for a fraction of the world's languages.

@@

\emph{Lexical selection} is the task of choosing, for a given SL word,
the most adequate translation in the TL among a known set of
alternatives.  The task is related to the task of word-sense
disambiguation \citep{ide98a}. However, it is different to word-sense
disambiguation in that lexical selection is a bilingual problem, not a
monolingual problem: its aim is to find the most adequate translation,
not the most adequate sense. Thus, it is not necessary to choose
between a series of fine-grained senses if all these senses result in
the same final translation; however, it may sometimes be necessary to
choose a different translation for the same sense, for example in a
collocation.
% It could also be the case that a single sense could have more than
% one possible translation (for example synonyms).

@@

Later, \cite{berger1996} illustrated the use of maximum-entropy
classifiers on the specific problem of lexical selection in IBM-style
word-based statistical MT. Other authors \citep{melero07a} have used
TL models to rank %@@@ carbonell06
the translations resulting from all possible combinations of lexical
selections. Nowadays, in state-of-the-art phrase-based
statistical % \todo{METIS too?}
MT \citep{koehn2009statistical}, lexical selection is taken care of by
a combination of the translation model and the language model. The
translation model provides probabilities of translation between words
or word sequences (often referred to as \emph{phrases}) in the source
and target language. The TL model provides probabilities of word
sequences in the TL.
% As regards rule-based machine translation, \cite{sanchez07a}
% proposed to use information from the target language to train a
% statistical lexical-selection module but did not provide any
% experiments.  @@@ \todo[inline]{FMT: integrate \citep{sanchez07a}
% somewhere here}
\cite{marechek10} trained a maximum-entropy lexical selector for their
dependency-grammar-based transfer system TectoMT using a bilingual
corpus.  More recently, \cite{tyers12a} presented a method of lexical
selection for RBMT based on rules which select or remove translations
in fixed-length contexts, along with a training method for learning
the rules from a word-aligned parallel corpus.

@@

\section{Method}
\todomlf{R3 says: ``The technique clearly resembles the "Deciphering
  Foreign Language," (S. Ravi and K. Knight) or “EM Decipherment for
  Large Vocabularies “ (M. Nuhn, and H. Ney. EM) approaches, although
  these techniques were not meant for learning the disambiguation but
  for learning the translation dictionaries.  The proposed method only
  executes one EM iteration in contrast to these works. It would be
  interesting to highlight differences and resemblances with these
  deciphering approaches.''. Check and cite? Also, single iteration
  WTF?}
%\todo[inline]{FSM: In the experiments a development corpus is
%  mentioned. For what is it used? This should be explain in this
%  section.}

@@

Let the probability of a word $t$ being the translation of a word $s$
in a SL context $c$ be $p_s(t|c)$. In principle, this value could be
estimated directly from the available corpora for every combination of
$(s, t, c)$.  This would however present two questions: (1) how should
the relevant contexts be chosen? and (2) what should be done when $(s,
t, c)$ is not found in the corpus?  A maximum-entropy model answers
both of these questions. It allows the contexts that we consider to be
linguistically interesting to be defined \emph{a priori} and then
integrate these seamlessly into a probabilistic model
\citep{Manning99b}.  In answer to the second question, a
maximum-entropy model maximise the entropy subject to match the
expected counts of the designed features with those found in the
training data. \todofsm{Solved R3: ``In answer to the second question, a
  maximum-entropy model assumes nothing about what is not present in
  the training data. That is, if there is no information in the
  training data, then it assumes that all outcomes —that is, all
  possible translations— are equally likely'' can be applied to a
  uniform distribution for a discrete variable. I think the authors
  are trying to restate the main property of the maximum entropy
  model, i.e.  maximum entropy model maximize the entropy (which is
  the uniformness of the distribution) subject to match the expected
  counts of the designed features with those found in the training
  data.  }
%assumes nothing about what is not present in the
%training data. 
That is, if there is no information in the training data, then it
assumes that all outcomes ---that is, all possible translations--- are
equally likely.  As previously mentioned, the principle of maximum
entropy has been applied to the problem of lexical selection before;
in particular, \cite{berger1996} cast the problem of lexical selection
in statistical MT as a classification problem. They learn a separate
maximum-entropy classifier for each SL word form,
%\comment{[:we do too, integrate this later]} 
using SL context to distinguish between possible translations. These
classifiers are then incorporated into the translation model of their
word-based statistical MT system.
%\comment{[how is our work different?]}  
In their approach, a classifier consists of a set of binary feature
functions and corresponding weights for each feature.  In both
\cite{berger1996} and our method, features are defined in the form
$h_k^{s}(t, c)$,\footnote{We follow the notation of \cite{berger1996}} where $t$ is a translation, and $c$ is a SL
context. One difference is that \cite{berger1996} take $s$, $t$ and
$c$ to be based on word forms, whereas in our method they are based on
lemma forms.  An example would be the following feature where the
Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' and would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_\mathrm{+handi}^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } 
\left\{ \begin{array}{c}
t~ =~ \textit{pez}~ \\ \mathrm{and}~ \\ \textit{handi}~ \mathrm{follows}~ \textit{arrain} \end{array} \right. \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word (+ \emph{handi}) to the right of it.

%%% THIS IS IN A WEIRD PLACE BUT WE WANT IT ON PAGE 3 -FMT
\begin{figure*}
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}
 \caption{A schema of the training process}
 \label{fig:lexselschema}
\end{figure*}

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of SL words, and $G = \{g_1, g_2, \ldots g_{|G|}\}$ is a set
of possible \emph{lexical-selection paths}. A lexical-selection path
$g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence of lexical-selection
choices of those SL words, where \(t_i\) is an element of \(T_s(s_i)\), the set of possible translations of \(s_i\).\todomlf{R1 asked for this, even if redundant, and says: ``can one source word be translated as multiple words? What about multi-word expressions that cannot be translated word-by-word or source words that do not have a TL equivalent'', etc.} % and \(T_s(\cdot)\)
%is the set of all
%translations for a source word.
This is produced in the first stages of RBMT, just after morphological
analysis, part-of-speech tagging, and bilingual dictionary lookup, and
before any structural transfer takes place (we will call this
\emph{pre-lexsel}). In our model, it is after these first stages that
lexical selection (\emph{lexsel}) occurs. After lexical selection,
structural transfer and generation take place; a function
$\tau(g_i,S)$ represents the result of these last stages, which we
will call \emph{post-lexsel}, and returns a finished translation of a
specific lexical-selection path $g_i$ of sentence
$S$. Figure~\ref{fig:lexselschema} shows this process
schematically. %%%\todo{Can we try to bring the figure closer?}
\todomlf{R3 says: ``it would have been nicer to see an example of one
  sentence going through the pipeline''. Can we add this in the text?
  FSM: We do not have much space}
%Taking our previous example for the parallel corpus, Table~\ref{table:ex-monocorp} 
%presents an example of five possible samples.

@@

suitable normalisation:\todomlf{R2 says ``Nevertheless, I've got
  problems to understand (4), in particular the paths in post-lexsel
  step. Some better explanation could be offered maybe by saying that
  (4) evaluates all possible solutions in the language model.''. Don't
  know what to do here. }

@@

\todomlf{R1 complains: ``the actual maximum entropy method itself is described very briefly (3 sentences at the end of Section 2)''}
\section{Experimental setting}

@@

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available
for evaluation (see Section~\ref{ss:measures}). 
% Table~\ref{table:lang-pair-stats} 
%presents statistics about the development status of each of the language 
%pairs chosen, including the number of bilingual dictionary entries, coverage of
%test corpora and translational ambiguity of the bilingual dictionary. Translation
%ambiguity is provided for both all words, and for only ambiguous open-category 
%words. In the experiments, lexical-selection rules are learnt for the open categories
%of adjective, noun and verb.

@@

\begin{description}
\addtolength{\itemsep}{-0.10in}

\item[Breton--French:] Development of the Breton--French pair has been
  described in \citep{tyers10b}. The bilingual % tyers09b
  dictionaries were not built with polysemy in mind from the outset,
  but some entries were added later to start work on lexical
  selection.\footnote{The version used in this paper is SVN revision
  41375; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-br-fr}}
%  This is the only available machine translation system between Breton
%  and French. It has been developed part-time over a number of years.
  \todomlf{R1 says ``whole section is rather length especially the
    revision numbers and URLs of the specific translations systems
    need not be mentioned in my opinion. The space could rather be
    used to elaborate in the description of the method in Sec
    2.''. FSM: I have moved them to the footnote}

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes
  parallel corpus \citep{tyers10}.
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this corpus
  using {\sc giza++} \citep{och03a} was checked to ensure that it was
  an adequate translation, and if so marked as the default.\footnote{A
    bilingual dictionary in Apertium \citep{forcada2011apertium} may
    contain several TL translation for a given SL word. Where there is
    more than one possible translation, the dictionary writer chooses
    the most general or most frequent translation among the set of
    possible translations and marks it as the \emph{linguistic
      default} translation.} As a result of attempting to include all
  possible translations, the average number of translations per word
  is much higher % (see table~\ref{table:lang-pair-stats})
  than in other pairs.\footnote{The version of the software used in this paper is SVN revision
  41476; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. 
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.\footnote{The version used is SVN revision 44846; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-eu-es}}
% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs, and
%  over a period of around four months. The pair
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with more
  than one translation.\footnote{The
  version used in this paper is SVN revision
  41387; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}

%The most-frequent translation, as judged by
%  the language pair developer, is marked as the \emph{default}
%  translation and non-default translations were added according to
%  frequency. \todo{Default is explained for some LPs only, why?}  
\todomlf{4-month mention removed in response to R1}
\end{description}

@@

The lexical-selection error rate is the fraction of times the given
system chooses a translation for a word which is not  the one found in an
annotated reference. The process uses a SL sentence, $S =
(s_1, s_2, \ldots, s_{|S|})$ and three functions. The first function,
$T_s(s_i)$, returns all possible translations of $s_i$ according to
the bilingual dictionary. The second function, $T_t(s_i)$, returns the
translations of $s_i$ selected by the lexical-selection module:
$T_t(s_i) \subseteq T_s(s_i)$; and usually $|T_t(s_i)| = 1$. If the
lexical-selection module returns more than one translation, the first
translation is selected. 
%! \comment{FSM: How many times does this
%!  happens?. I would move this last sentence to an appendix.}  
The
function $T_r(s_i)$ returns the set of reference translations which
are acceptable for $s_i$ in sentence $S$.\footnote{Depending on how
  the reference is built, the set returned by $T_r(s_i)$ % may be
  % incomplete in the sense that it 
  may not include all possible
  acceptable translations.} For a
single sentence, we define the lexical selection error rate (LER) of
that sentence as\todomlf{R3 says: ``The authors use LER, but F-score measures seem more appropriate to estimate the performance of the disambiguation module. Moreover, the constraint of the disambiguation module to generate only one word when evaluating seems extreme.'' He don't get it.}
\begin{equation}
\mathrm{LER} = \frac{\sum_{i=1}^{|S|} \mathrm{amb}(s_i)\; \mathrm{diff}(T_r(s_i), T_t(s_i))}{\sum_{i=1}^{|S|} \mathrm{amb}(s_i)},
\label{eq:ler}
\end{equation}
where
\begin{equation}
\mathrm{amb}(s_i) = \left \{ \begin{matrix}
                 1 & \mbox{if } |T_s(s_i)| > 1   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:amb}
\end{equation}
 tests if a word is ambiguous, and the  function 
\begin{equation}
\mathrm{diff}(T_r(s_i), T_t(s_i)) = \left \{ \begin{matrix}
                 1 & \mbox{if } T_r(s_i) \cap T_t(s_i) = \emptyset   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:diff}
\end{equation}
states that there is a difference if the intersection between the set
of reference translations $T_r(s_i)$ and the set of translations from
the lexical selection module $T_t(s_i)$ is empty.  Recall
that,\todomlf{R1 says ``this has already been mentioned
  before''. Don't know what to do FSM: I would do nothing, this is why
the sentence says ``Recall ...''} although $T_t(s_i)$ returns a set,
this set will be a singleton, as when the lexical-selection module
returns more than one translation the first one will be
selected.\footnote{In practice this does not happen as each ambiguous
  word has a \emph{default} translation.}\todomlf{R1 complains: ``why
  mention if it doesn't happen in practice? Relevance?'' What should
  we do?}

@@

The table in Figure~\ref{fig:ler-example} gives an
overview of the inputs.\todomlf{Caption of figure extended in response to R3} In the description it is assumed that the
reference translation has been annotated by hand. However, hand
annotation is a time-consuming process, and was not possible. A
description of how the reference was built is given in
Section~\ref{sec:eval-corpus}.

>>>>

\begin{abstract}
  This article presents a method of training maximum-entropy models to
  perform lexical selection in a rule-based machine translation
  system. The training method described is unsupervised; that is, it
  does not require any annotated corpus. The method uses
  source-language monolingual corpora, the machine translation (MT)
  system in which the models are integrated, and a statistical
  target-language model. Using the MT system, the sentences in 
  the source-language corpus are
  translated in all possible ways according to the different
  translation equivalents in the bilingual dictionary of the system.
  These translations are then scored on the target-language model and
  the scores are normalised to provide fractional counts for training
  source-language maximum-entropy lexical-selection models. We show
  that these models can perform equally well, or better, than using
  the target-language model directly for lexical selection, at a
  substantially reduced computational cost.
\end{abstract}

\section{Introduction}
%\todomlf{R1 has a general comment: ``watch out for long sentences''}
Corpus-based machine translation (MT) has been the primary research
direction in the field of MT in recent years. However, rule-based MT
(RBMT) systems are still being developed, and there are many
successful commercial and non-commercial systems. One reason for the
continued development of RBMT systems is that in order to be
successful, corpus-based MT requires parallel corpora in the order of
tens of millions of words. Although for some language pairs these
exist, they only exist for a fraction of the world's languages.

@@

\emph{Lexical selection} is the task of choosing, for a given SL word,
the most adequate translation in the TL among a known set of
alternatives.  The task is related to the task of word-sense
disambiguation \citep{ide98a}. However, it is different to word-sense
disambiguation in that lexical selection is a bilingual problem, not a
monolingual problem: its aim is to find the most adequate translation,
not the most adequate sense. Thus, it is not necessary to choose
among a series of fine-grained senses if all these senses result in
the same final translation; however, it may sometimes be necessary to
choose a different translation for the same sense, for example in a
collocation.
% It could also be the case that a single sense could have more than
% one possible translation (for example synonyms).

@@

Later, \cite{berger1996} illustrated the use of maximum-entropy
classifiers on the specific problem of lexical selection in IBM-style
word-based statistical MT. Other authors \citep{melero07a} have used
TL models to rank %@@@ carbonell06
the translations resulting from all possible combinations of lexical
selections. Nowadays, in state-of-the-art phrase-based
statistical % \todo{METIS too?}
MT \citep{koehn2009statistical}, lexical selection is taken care of by
a combination of the translation model and the language model. The
translation model provides probabilities of translation between words
or word sequences (often referred to as \emph{phrases}) in the source
and target language. The TL model provides probabilities of word
sequences in the TL.
% As regards rule-based machine translation, \cite{sanchez07a}
% proposed to use information from the target language to train a
% statistical lexical-selection module but did not provide any
% experiments.  @@@ \todo[inline]{FMT: integrate \citep{sanchez07a}
% somewhere here}
\cite{marechek10} trained a maximum-entropy lexical selector for their
dependency-grammar-based transfer system TectoMT using a bilingual
corpus.  More recently, \cite{tyers12a} presented a method of lexical
selection for RBMT based on rules which select or remove translations
in fixed-length contexts, along with a training method for learning
the rules from a word-aligned parallel corpus.\footnote{The work by \cite{ravi2011deciphering} and \cite{nuhn2014decipherment}, who decipher word-ciphered text using monolingual corpora only may be seen as a generalised version of the problem of lexical selection without parallel corpora.}

@@

\section{Method}
% \todomlf{R3 says: ``The technique clearly resembles the "Deciphering
%   Foreign Language," (S. Ravi and K. Knight) or “EM Decipherment for
%   Large Vocabularies “ (M. Nuhn, and H. Ney. EM) approaches, although
%   these techniques were not meant for learning the disambiguation but
%   for learning the translation dictionaries.  The proposed method only
%   executes one EM iteration in contrast to these works. It would be
%   interesting to highlight differences and resemblances with these
%   deciphering approaches.''. Check and cite? Also, single iteration
%   WTF?}
%\todo[inline]{FSM: In the experiments a development corpus is
%  mentioned. For what is it used? This should be explain in this
%  section.}

@@

Let the probability of a word $t$ being the translation of a word $s$
in a SL context $c$ be $p_s(t|c)$. In principle, this value could be
estimated directly from the available corpora for every combination of
$(s, t, c)$.  This would however present two questions: (1) how should
the relevant contexts be chosen? and (2) what should be done when $(s,
t, c)$ is not found in the corpus?  A maximum-entropy model answers
both of these questions. It allows the contexts that we consider to be
linguistically interesting to be defined \emph{a priori} and then
integrate these seamlessly into a probabilistic model
\citep{Manning99b}.  In answer to the second question, a
maximum-entropy model maximises the entropy subject to match the
expected counts of the designed features with those found in the
training data. % \todofsm{Solved R3: ``In answer to the second question, a
  % maximum-entropy model assumes nothing about what is not present in
  % the training data. That is, if there is no information in the
  % training data, then it assumes that all outcomes —that is, all
  % possible translations— are equally likely'' can be applied to a
  % uniform distribution for a discrete variable. I think the authors
  % are trying to restate the main property of the maximum entropy
  % model, i.e.  maximum entropy model maximize the entropy (which is
  % the uniformness of the distribution) subject to match the expected
  % counts of the designed features with those found in the training
  % data.  }
%assumes nothing about what is not present in the
%training data. 
That is, if there is no information in the training data, then it
assumes that all outcomes ---that is, all possible translations--- are
equally likely.  As previously mentioned, the principle of maximum
entropy has been applied to the problem of lexical selection before;
in particular, \cite{berger1996} cast the problem of lexical selection
in statistical MT as a classification problem. They learn a separate
maximum-entropy classifier for each SL word form,
%\comment{[:we do too, integrate this later]} 
using SL context to distinguish between possible translations. These
classifiers are then incorporated into the translation model of their
word-based statistical MT system.
%\comment{[how is our work different?]}  
In their approach, a classifier consists of a set of binary feature
functions and corresponding weights for each feature.  In both
\cite{berger1996} and our method, features are defined in the form
$h_k^{s}(t, c)$,\footnote{We follow the notation of \cite{berger1996}} where $t$ is a translation, and $c$ is a SL
context. One difference is that \cite{berger1996} take $s$, $t$ and
$c$ to be based on word forms, whereas in our method they are based on
lemma forms.  An example would be the following feature where the
Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' and would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_\mathrm{+handi}^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } 
\left\{ \begin{array}{c}
t~ =~ \textit{pez}~ \\ \mathrm{and}~ \\ \textit{handi}~ \mathrm{follows}~ \textit{arrain} \end{array} \right. \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word (+ \emph{handi}) to the right of it.

%%% THIS IS IN A WEIRD PLACE BUT WE WANT IT ON PAGE 3 -FMT
\begin{figure*}
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.9cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{0.9cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.9cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}
 \caption{A schema of the lexical selection process: source sentence \(S\) has \(|G|\) lexical selection paths \(g_i\): \emph{lexsel} selects one of them \(g^\star\), which is used to generate translation \(\tau(g^\star,S)\).}
 \label{fig:lexselschema}
\end{figure*}

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of SL words, and $G = \{g_1, g_2, \ldots g_{|G|}\}$ is a set
of possible \emph{lexical-selection paths}. A lexical-selection path
$g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence of lexical-selection
choices of those SL words, where \(t_i\) is an element of \(T_s(s_i)\), the set of possible translations of \(s_i\).\footnote{We deal only with single-word translations in this paper.}
% \todomlf{R1 asked for this, even if redundant, and says: ``can one source word be translated as multiple words? What about multi-word expressions that cannot be translated word-by-word or source words that do not have a TL equivalent'', etc.} % and \(T_s(\cdot)\)
%is the set of all
%translations for a source word.
This is produced in the first stages of RBMT, just after morphological
analysis, part-of-speech tagging, and bilingual dictionary lookup, and
before any structural transfer takes place (we will call this
\emph{pre-lexsel}). In our model, it is after these first stages that
lexical selection (\emph{lexsel}) occurs. After lexical selection,
structural transfer and generation take place; a function
$\tau(g_i,S)$ represents the result of these last stages, which we
will call \emph{post-lexsel}, and returns a finished translation of a
specific lexical-selection path $g_i$ of sentence
$S$. Figure~\ref{fig:lexselschema} shows this process
schematically. %%%\todo{Can we try to bring the figure closer?}
% \todomlf{R3 says: ``it would have been nicer to see an example of one
%   sentence going through the pipeline''. Can we add this in the text?
%   FSM: We do not have much space}
%Taking our previous example for the parallel corpus, Table~\ref{table:ex-monocorp} 
%presents an example of five possible samples.

@@

suitable normalisation:%\todomlf{R2 says ``Nevertheless, I've got
  % problems to understand (4), in particular the paths in post-lexsel
  % step. Some better explanation could be offered maybe by saying that
  % (4) evaluates all possible solutions in the language model.''. Don't
  % know what to do here. }

@@

% \todomlf{R1 complains: ``the actual maximum entropy method itself is described very briefly (3 sentences at the end of Section 2)''}
\section{Experimental setting}

@@

Evaluation will be performed using four Apertium \citep{forcada2011apertium} language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available
for evaluation (see Section~\ref{ss:measures}).\footnote{The Apertium revision (version) used is given in footnotes.}
% Table~\ref{table:lang-pair-stats} 
%presents statistics about the development status of each of the language 
%pairs chosen, including the number of bilingual dictionary entries, coverage of
%test corpora and translational ambiguity of the bilingual dictionary. Translation
%ambiguity is provided for both all words, and for only ambiguous open-category 
%words. In the experiments, lexical-selection rules are learnt for the open categories
%of adjective, noun and verb.

@@

\begin{description}
\addtolength{\itemsep}{-1ex}
\item[Breton--French]  \citep{tyers10b}: Bilingual
  dictionaries were not built with polysemy in mind from the outset,
  but some entries were added later to start work on lexical
  selection.\footnote{Revision
  41375; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-br-fr}}
%  This is the only available machine translation system between Breton
%  and French. It has been developed part-time over a number of years.
\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes
  parallel corpus \citep{tyers10}.
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this corpus
  using {\sc giza++} \citep{och03a} was checked to ensure that it was
  an adequate translation, and if so marked as the
  default.\footnote{Bilingual dictionaries in Apertium
    \citep{forcada2011apertium} may contain several translations for a
    given word. Dictionary writers may mark as \emph{linguistic default} the most
    general or most frequent translation among the set of possible
    translations.} As a result of attempting to include all possible
  translations, the average number of translations per word is much
  higher % (see table~\ref{table:lang-pair-stats})
  than in other pairs.\footnote{Revision 41476;
    \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}
\item[Basque--Spanish]  \citep{ginesti09}:
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  alternative translations were included in the bilingual
  dictionary.\footnote{Revision 44846; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-eu-es}}
% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs, and
%  over a period of around four months. The pair
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with more
  than one translation.\footnote{Revision
  41387; \url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
%The most-frequent translation, as judged by
%  the language pair developer, is marked as the \emph{default}
%  translation and non-default translations were added according to
%  frequency. \todo{Default is explained for some LPs only, why?}  
\end{description}

@@

The lexical-selection error rate is the fraction of times the given
system chooses a translation for a word which is not  the one found in an
annotated reference. The process uses a SL sentence, $S =
(s_1, s_2, \ldots, s_{|S|})$ and three functions. The first function,
$T_s(s_i)$, returns all possible translations of $s_i$ according to
the bilingual dictionary. The second function, $T_t(s_i)$, returns the
translations of $s_i$ selected by the lexical-selection module:
$T_t(s_i) \subseteq T_s(s_i)$; and usually $|T_t(s_i)| = 1$. If the
lexical-selection module returns more than one translation, the first
translation is selected. 
%! \comment{FSM: How many times does this
%!  happens?. I would move this last sentence to an appendix.}  
The
function $T_r(s_i)$ returns the set of reference translations which
are acceptable for $s_i$ in sentence $S$.\footnote{Depending on how
  the reference is built, the set returned by $T_r(s_i)$ % may be
  % incomplete in the sense that it 
  may not include all possible
  acceptable translations.} For a
single sentence, we define the lexical selection error rate (LER) of
that sentence as% \todomlf{R3 says: ``The authors use LER, but F-score measures seem more appropriate to estimate the performance of the disambiguation module. Moreover, the constraint of the disambiguation module to generate only one word when evaluating seems extreme.'' He don't get it.}
\begin{equation}
\mathrm{LER} = \frac{\sum_{i=1}^{|S|} \mathrm{amb}(s_i)\; \mathrm{diff}(T_r(s_i), T_t(s_i))}{\sum_{i=1}^{|S|} \mathrm{amb}(s_i)},
\label{eq:ler}
\end{equation}
where
\begin{equation}
\mathrm{amb}(s_i) = \left \{ \begin{matrix}
                 1 & \mbox{if } |T_s(s_i)| > 1   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:amb}
\end{equation}
 tests if a word is ambiguous, and the  function 
\begin{equation}
\mathrm{diff}(T_r(s_i), T_t(s_i)) = \left \{ \begin{matrix}
                 1 & \mbox{if } T_r(s_i) \cap T_t(s_i) = \emptyset   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:diff}
\end{equation}
states that there is a difference if the intersection between the set
of reference translations $T_r(s_i)$ and the set of translations from
the lexical selection module $T_t(s_i)$ is empty.  Recall
that, %\todomlf{R1 says ``this has already been mentioned
%  before''. Don't know what to do FSM: I would do nothing, this is why
%the sentence says ``Recall ...''} 
although $T_t(s_i)$ returns a set,
this set will be a singleton, as when the lexical-selection module
returns more than one translation, Apertium will select the default one if marked or the first one of not.\footnote{In practice this does not happen as each ambiguous
  word has a \emph{default} translation.}%\todomlf{R1 complains: ``why
  % mention if it doesn't happen in practice? Relevance?'' What should
  % we do?}

@@

The table in Figure~\ref{fig:ler-example} gives an
overview of the inputs.%\todomlf{Caption of figure extended in response to R3} 
In the description it is assumed that the
reference translation has been annotated by hand. However, hand
annotation is a time-consuming process, and was not possible. A
description of how the reference was built is given in
Section~\ref{sec:eval-corpus}.

\subsubsection{Machine translation performance}
